---
title: "Lab 11 HW"
author: "Lorena Cuellar"
date: "2022-11-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

The sample size for each genotype
```{r}
table(expr$geno)
```
Median expression levels for each of these genotypes
```{r}
library(doBy)
summaryBy(exp~geno, data = expr, FUN=median)
```

```{r}
summary(expr)
```

```{r}
library(ggplot2)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

It can be inferred that the the A/A genotype has higher expression levels than the G/G genotype from this plot. Additionally, it does seem like this SNP effects the expression of ORMDL3, it looks like the SNP increases ORMDL3 expression.

```{r}
ggplot(expr) + aes(x=geno, exp, fill=geno) + geom_boxplot(notch=TRUE)
```


