---
title: "Lab 10: Halloween Mini-Project"
author: Lorena Cuellar
format: gfm
---


```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```


> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)

```



```{r}
candy["Twix", ]$winpercent
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Caramel Apple Pops", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```


```{r}
library("skimr")
skim(candy)
```


```{r}
skimr::skim(candy)
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

A: The winpercent variable seems to be on a different scales compared to the other variables which in their columns they seem to have mostly 0 to 1 values.


>Q7. What do you think a zero and one represent for the candy$chocolate column?

A: I think the 0 and 1 represents if the candy has chocolate or not. So a 1 would indicate yes and a 0 would indicate no.

```{r}
candy$chocolate
skim(candy$chocolate)
```

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy) + aes(winpercent) + geom_histogram(bins=10, col="brown", fill="dark green")
```


>Q9. Is the distribution of winpercent values symmetrical?

A: No it is not symmetrical, the values seem to be slanted on one side.


>Q10. Is the center of the distribution above or below 50%?

A: The center is below 50%

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate.inds <- as.logical(candy$chocolate)
choco <- candy[chocolate.inds,]$winpercent
mean(choco)


fruity.inds <- as.logical(candy$fruity)
fruitys <-  candy[fruity.inds,]$winpercent
mean(fruitys)
```

A: Chocolate candy is ranked higher than fruity candy

> Q12. Is this difference statistically significant?

A: It is significant, people prefer chocolate.

```{r}
t.test(choco, fruitys)
```

# Candy Ranking

>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5)
```




> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col() 
```


```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"


```


```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```


```{r}
ggsave("tmp.png")
```

> Q17. What is the worst ranked chocolate candy?

A: Nik L Nip

>Q18. What is the best ranked fruity candy?

A: Reeses Peanut Butter cup

# 4: Taking a look at Pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```



> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

A: Reesses Miniatures if you like candy but also sour patch kids if you like fruity candy.

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

A: The least popular of the top 5 most expensive is the Nik L Nip candy


```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()

```

# 5 Correlation Strucutre

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

A: Chocolate and fruity


>Q23. Similarly, what two variables are most positively correlated?

A: chocolate and bar


# 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Plot our main PCA

```{r}
plot(pca$x[,1], pca$x[,2])
```

Add some color
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```


Make a nicer plot

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

A: The variables that are picked up the strongest by PC1 are fruity, hard, and pluribus. These make sense because if we look at plots above we can see that these are usually corelated with each other so it makes sense why they have positive direction in PC1.